
##HeLa 

```{r, include= False}
library(readr)
library(dplyr)
library(tidyverse)
```


```{r}
## first line is causing erros, so lets skip it and name the columns later
raw <- read_delim(
  "../data/GSE247417_raw_counts.txt",
  delim = " ",        
  skip = 1,           
  col_names = FALSE,  
  trim_ws = TRUE,
  show_col_types = FALSE
)
```



```{r}
colnames(raw) <- c(
  "gene_id",
  "GFP_rep1", "GFP_rep2", "GFP_rep3",
  "HS-GFP_rep1", "HS-GFP_rep2", "HS-GFP_rep3",
  "WT_rep1", "WT_rep2", "WT_rep3",
  "HS-WT_rep1", "HS-WT_rep2", "HS-WT_rep3"
)


dim(raw)
head(raw)


```

	•	GFP vs HS-GFP = “normal HeLa cells under heat shock vs control”
	•	WT vs HS-WT = “HeLa cells with extra CLK2 under heat shock vs control”

we don’t want to mix in CLK2 overexpression effects, so we just use GFP vs HS-GFP and ignore the “WT” 

```{r}
human_clean <- raw %>%
  dplyr::select(
    gene_id,
    GFP_rep1, GFP_rep2, GFP_rep3,
    `HS-GFP_rep1`, `HS-GFP_rep2`, `HS-GFP_rep3`
  )

dim(human_clean)
head(human_clean)

```


```{r}
write_csv(human_clean, "../data/human_GSE247417_clean_counts.csv")
```


## Mouse HS 

```{r}
mouse <- read_csv("../data/GSE133251_gene_fpkm_matrix.csv.gz")

colnames(mouse)
dim(mouse)
head(mouse)
```
The columns mean the following:
  • WT_NT_* = wild-type YAP/TAZ normal (no knockdown), no treatment
	•	WT_HS_* = wild-type YAP/TAZ, heat shock
	•	YAPTAZDKd_NT_* = YAP/TAZ double knockdown, no treatment
	•	YAPTAZDKd_HS_* = YAP/TAZ double knockdown + heat shock
	
For the project we only need:
	•	WT_NT_1–3 = control
	•	WT_HS_1 & WT_HS_3 = heat shock


```{r}
library(tidyverse)


mouse_clean <- mouse %>%
  transmute(
    ensembl_id = `#GeneID`,
    WT_NT_1    = FPKM.WT_NT_1,
    WT_NT_2    = FPKM.WT_NT_2,
    WT_NT_3    = FPKM.WT_NT_3,
    WT_HS_1    = FPKM.WT_HS_1,
    WT_HS_3    = FPKM.WT_HS_3
  )

dim(mouse_clean)
head(mouse_clean)
```

```{r}
write_csv(mouse_clean, "mouse_WT_heatshock_FPKM_clean.csv")
```

## Fibroblasts

```{r}


human_raw <- read_tsv(
  "../data/GSE100469_FPKM_stress.txt",
)

colnames(human_raw)[1] <- "gene_id"

human_raw <- human_raw %>%
  dplyr::filter(startsWith(gene_id, "ENSG"))

human_mock_heat1h <- human_raw %>%
  dplyr::select(
    gene_id,
    mock1_1, mock2_1,    # control, replicate 1
    mock1_2, mock2_2,    # control, replicate 2
    heat_1h_1,           # 1h HS, repl 1
    heat_1h_2            # 1h HS, repl 2
  )


dim(human_mock_heat1h)
head(human_mock_heat1h)
```

```{r}

write_csv(human_mock_heat1h, "../data/GSE100469_FPKM_human_Fibroblast_HS_Clean.csv")
```