

```{r, include=FALSE}
library(limma)
library(tidyverse)
library(biomaRt)
library(dplyr)
library(EnsDb.Hsapiens.v86)
library(ensembldb)
```
# Limma

## Fibroblasts

```{r}
fib <- read.csv("../data/GSE100469_FPKM_human_Fibroblast_HS_Clean.csv")
```


```{r}
expr_raw <- fib %>%
 
  column_to_rownames("gene_id") %>%
  as.matrix()
expr_log <- log2(expr_raw + 1)


expr_log <- normalizeBetweenArrays(expr_log, method = "quantile")
```



```{r}
samples <- colnames(expr_log)


coldata <- tibble(
  sample = samples,
  condition = factor(
    c("control", "control", "control", "control", "heatshock", "heatshock"),
    levels = c("control", "heatshock")
  )
)
design
coldata$condition <- factor(coldata$condition, levels = c("control", "heatshock"))
design <- model.matrix(~ condition, data = coldata)

fit <- lmFit(expr_log, design)
fit <- eBayes(fit, trend = TRUE)

res_fib <- topTable(fit, coef = "conditionheatshock", number = Inf, sort.by = "P")
```

```{r}
head(res_fib)
```

```{r}
deg_limma <- res_fib %>% rownames_to_column("gene_id") %>%
  dplyr::filter(!is.na(adj.P.Val), adj.P.Val < 0.05, abs(logFC) > 0.5)
```

```{r}
nrow(deg_limma)
```


```{r}
write.csv(deg_limma, "../data/Fibroblasten_DEGs.csv")
```


## Mouse


```{r}
mouse <- read_csv("../data/mouse_WT_heatshock_FPKM_clean.csv")

head(mouse)

```

Limma assumes a normal dist. , so we apply a log tranformation

```{r}

expr_raw <- mouse %>%
  column_to_rownames("ensembl_id") %>%
  as.matrix()

expr_log <- log2(expr_raw + 1)

```

defining sample metadata:
control 3 NT and 2 HS
```{r}
samples <- colnames(expr_log)

coldata <- tibble(
  sample = samples,
  condition = if_else(str_detect(sample, "WT_HS"), "heatshock", "control")
) %>%
  mutate(condition = factor(condition, levels = c("control", "heatshock")))

table(coldata$condition)


expr_log <- expr_log[, coldata$sample, drop = FALSE]

```
 design matrices tell which group which sample belongs to (whether control or heatshock)

```{r}

design <- model.matrix(~ 0 + condition, data = coldata)
colnames(design) <- levels(coldata$condition)
design
```

first we fit the data, we create a contrast which describe the comparison we want to look at(here heatshock vs control). Then we apply the contrast and use eBayes() to estimate variance for each gene

```{r}
fit <- lmFit(expr_log, design)

contrast <- makeContrasts(
  HS_vs_CTRL = heatshock - control,
  levels = design
)

fit2 <- contrasts.fit(fit, contrast)
fit2 <- eBayes(fit2, trend = TRUE)

```


```{r}
res_mouse <- topTable(
  fit2,
  coef = "HS_vs_CTRL",
  number = Inf,
  sort.by = "P"
)

head(res_mouse)

summary(res_mouse$logFC)
summary(-log10(res_mouse$P.Value))

```


```{r}
deg_mouse <- res_mouse %>%
  rownames_to_column("ensembl_id") %>%
  dplyr::filter(adj.P.Val < 0.05, abs(logFC) > 0.5)

```



```{r}
nrow(deg_mouse)
```

```{r}
write.csv(deg_mouse, "../data/mouse_DEGs.csv", row.names = FALSE)
```

## looking at log 2 fold change spread between data sets 

```{r}




hela_deg <- read.csv("../data/HeLa_DEGs.csv")          
fib_deg  <- read.csv("../data/Fibroblasten_DEGs.csv") 
mouse_deg<- read.csv("../data/mouse_DEGs.csv")        
comp <- tibble(
  dataset = c(rep("HeLa (DESeq2)", nrow(hela_deg)),
              rep("Fibroblast (limma)", nrow(fib_deg)),
              rep("Mouse (limma)", nrow(mouse_deg))),
  lfc = c(hela_deg$log2FoldChange, fib_deg$logFC, mouse_deg$logFC)
)

comp %>%
  group_by(dataset) %>%
  summarise(
    n = n(),
    lfc_IQR = IQR(lfc, na.rm=TRUE),
    lfc_q05 = quantile(lfc, 0.05, na.rm=TRUE),
    lfc_q95 = quantile(lfc, 0.95, na.rm=TRUE)
  )

ggplot(comp, aes(x=dataset, y=lfc)) +
  geom_boxplot(outlier.alpha = 0.2) +
  theme_minimal() +
  labs(x=NULL, y="log2 fold change (DESeq2) / logFC (limma)")
```
```




